\section{Ablation Studies}

We conduct two ablation studies on ConvNeXt-Tiny to isolate the effects of optimizer choice and data augmentation. Both use the same hyperparameters as the main experiments but with reduced training budgets.

\subsection{Optimizer: AdamW vs.\ SGD}

We compare AdamW and SGD with momentum (0.9) over 10 epochs. Learning rates are tuned independently for a fair comparison: AdamW uses $\eta_{\text{bb}}{=}10^{-4}$, $\eta_{\text{head}}{=}10^{-3}$; SGD uses $\eta_{\text{bb}}{=}5{\times}10^{-3}$, $\eta_{\text{head}}{=}5{\times}10^{-2}$. As shown in Tab.~\ref{tab:ablation_opt}, AdamW achieves 97.81\% validation accuracy compared to 96.23\% for SGD, a gap of 1.58 pp. AdamW also converges faster, reaching 95.40\% by epoch 2, while SGD requires epoch 4 to surpass 95\%. This advantage is consistent with prior findings that adaptive optimizers are better suited for fine-tuning pretrained models, where different parameter groups benefit from per-parameter learning rate adaptation~\citep{loshchilov2019decoupled}.

\begin{table}[t]
\centering
\caption{\hl{Ablation: Optimizer} (ConvNeXt-Tiny, 10 epochs). Learning rates are tuned independently per optimizer.}
\label{tab:ablation_opt}
\small
\begin{tabular}{lccc}
\toprule
Optimizer & LR (bb / head) & Best Val Acc (\%) & Final Train Loss \\
\midrule
AdamW & $10^{-4}$ / $10^{-3}$                       & \cellcolor{blue!8}97.81 & 0.794 \\
SGD   & $5{\times}10^{-3}$ / $5{\times}10^{-2}$     & 96.23 & 0.865 \\
\bottomrule
\end{tabular}
\end{table}

Tab.~\ref{tab:ablation_opt_epochs} tracks epoch-wise convergence. AdamW jumps from 21.87\% to 95.40\% between epochs 1 and 2, then steadily improves. SGD shows a more gradual trajectory, reaching 88.39\% at epoch 2 and continuing to climb through epoch 10. Both optimizers converge to competitive accuracies, but AdamW's faster early convergence makes it preferable under limited training budgets.

\begin{table}[t]
\centering
\caption{\hl{Optimizer convergence.} Val accuracy (\%) per epoch.}
\label{tab:ablation_opt_epochs}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccccccc}
\toprule
 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\
\midrule
AdamW & 21.9 & 95.4 & 96.4 & 96.6 & 96.0 & 96.6 & 96.6 & 97.1 & 97.5 & \cellcolor{blue!8}97.8 \\
SGD   & 24.4 & 88.4 & 93.8 & 95.0 & 95.3 & 95.6 & 95.9 & 96.2 & 96.2 & \cellcolor{blue!8}96.2 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Data Augmentation}

We train ConvNeXt-Tiny for 20 epochs with and without data augmentation. The ``with'' variant applies random crop, horizontal flip, color jitter, and random erasing ($p{=}0.25$); the ``without'' variant uses only resize, center crop, and ImageNet normalization. As shown in Tab.~\ref{tab:ablation_aug}, both variants achieve similar validation accuracy (97.51\% vs.\ 97.74\%), and the train--validation gaps are comparable (2.53\% vs.\ 2.44\%). Both reach near-perfect training accuracy ($>$99.8\%) by epoch 6--7.

Contrary to the typical expectation that augmentation prevents overfitting, removing augmentation does not degrade performance. This can be attributed to the strong inductive bias of ImageNet-pretrained features: the pretrained backbone already captures robust, augmentation-invariant representations, reducing the marginal benefit of explicit augmentation on a moderately sized dataset like Caltech-101 with relatively low intra-class variation.

\begin{table}[t]
\centering
\caption{\hl{Ablation: Data augmentation} (ConvNeXt-Tiny, 20 epochs).}
\label{tab:ablation_aug}
\small
\begin{tabular}{lcccc}
\toprule
Augmentation & Best Val Acc (\%) & Final Train Acc (\%) & Train--Val Gap & Final Train Loss \\
\midrule
With    & 97.51 & 99.89 & 2.53\% & 0.787 \\
Without & \cellcolor{blue!8}97.74 & 99.95 & 2.44\% & 0.782 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Summary}

Tab.~\ref{tab:ablation_summary} summarizes both ablations. Optimizer choice has a meaningful impact: AdamW outperforms SGD by 1.58 pp under matched training budgets, with notably faster convergence. Data augmentation, on the other hand, provides no measurable benefit when fine-tuning a strong pretrained model on Caltech-101, suggesting that the learned ImageNet representations are already sufficiently robust for this task.

\begin{table}[t]
\centering
\caption{\hl{Ablation summary.} Best validation accuracy (\%) on ConvNeXt-Tiny.}
\label{tab:ablation_summary}
\small
\begin{tabular}{llccc}
\toprule
Factor & Setting & Val Acc (\%) & $\Delta$ (pp) \\
\midrule
\multirow{2}{*}{Optimizer}    & SGD (tuned LR)   & 96.23 & -- \\
                               & AdamW             & \cellcolor{blue!8}97.81 & +1.58 \\
\addlinespace
\multirow{2}{*}{Augmentation} & Without           & \cellcolor{blue!8}97.74 & -- \\
                               & With              & 97.51 & $-$0.23 \\
\bottomrule
\end{tabular}
\end{table}