\section{Related Work}

\textbf{CNN Architectures.} Since \texttt{AlexNet}~\citep{krizhevsky2012imagenet}, convolutional networks have been the backbone of image recognition. \texttt{ResNet}~\citep{he2016deep} enabled much deeper networks through residual connections, while \texttt{EfficientNet}~\citep{tan2019efficientnet} systematically scaled depth, width, and resolution to improve the accuracy--efficiency trade-off. \texttt{ConvNeXt}~\citep{liu2022convnet} later incorporated design cues from Transformers (e.g., larger kernels, layer normalisation) while retaining a purely convolutional structure. These architectures are routinely pretrained on ImageNet and transferred to downstream tasks including medical image analysis.

\textbf{Vision Transformers and Self-supervised Pre-training.} \texttt{ViT}~\citep{dosovitskiy2020image} showed that pure self-attention over image patches can match or surpass CNNs given sufficient data. Subsequent work reduced the data requirement through stronger pre-training: \texttt{EVA-02}~\citep{fang2024eva} combines masked image modelling with CLIP-guided distillation to learn transferable representations from unlabelled images. Such self-supervised approaches are particularly promising for medical AI, where large labelled datasets are scarce.

\textbf{Feature-based Classifiers.} Before end-to-end deep learning, handcrafted features paired with SVMs were standard for object recognition~\citep{fei2004learning}. The hybrid strategy of extracting features from a frozen pretrained CNN and training a lightweight classifier on top~\citep{donahue2014decaf} remains practical when compute or labelled data is limited, a common constraint in clinical deployments.

\textbf{Benchmarking on Medium-scale Datasets.} Caltech-101~\citep{fei2004learning} has long served as a testbed for comparing classification methods, yet most prior studies evaluate a single architecture or use inconsistent data splits, making fair comparison difficult. Our work addresses this by benchmarking six methods under a unified protocol on the same stratified split.